---
title: "Creativity-AI"
author: "Claudia Alvarez-Martin, Dan Dediu"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: yes
    toc_depth: 6
    toc_float:
      collapsed: yes
      smooth_scroll: no
    highlight: tango
    theme: cerulean
    code_folding: hide
editor_options: 
  chunk_output_type: console
knit: (function(input, ...) {
    )
    rmarkdown::render(
      input,
      output_dir = "../results"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE)

# use the project folder as root directory
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```


```{r libraries and initializations, results='hide', warning=FALSE, message=FALSE}
# Rmarkdown:
library(knitr);
library(pander);

# Load and write .xlsx data:
library(readxl);
library(openxlsx)

# Manage data;
library(tidyr);
library(dplyr);

# Plotting:
library(reshape2);
library(dplyr);
library(ggplot2);
library(gridExtra);
library(sjPlot);
library(viridis);
library(ggsignif)
library(ggpubr)
library(ggrepel)

# Psychometrics:
library(performance);
library(factoextra);
library(NbClust);
library(mokken);
library(psych);
library(lavaan);
library(lavaanPlot);

# IRR
library(irr)

# Regression:
library(lmerTest);
library(glmmTMB);
#library(brms);
library(effects);
library(emmeans);

# System/parallel processing:
library(parallel);

# Misc:
library(benchmarkme);

if(!dir.exists("./cache")) dir.create("./cache", showWarnings=FALSE, recursive=TRUE);

# For parallel processing:
n_cores <- ifelse(Sys.info()['sysname'] == "Windows", 1, max(detectCores(all.tests=TRUE, logical=FALSE)-1, 1, na.rm=TRUE)); # try to use multiple cores, if running on a non-Windows system

## Set seed for reproducibility:
set.seed(42);

# Figure and Table caption adapted from https://stackoverflow.com/questions/37116632/rmarkdown-html-number-figures: 
outputFormat = opts_knit$get("rmarkdown.pandoc.to"); # determine the output format of the document
if( is.null(outputFormat) ) outputFormat = ""; # probably not run within knittr
capTabNo = 1; capFigNo = 1; # figure and table caption numbering, for HTML do it manually
#Function to add the Table Number
capTab = function(x, supplementary=FALSE, id=TRUE)
{
  if(outputFormat == 'html'){
    #x = paste0("<div class='caption'>**Table ",ifelse(supplementary,"S",""),capTabNo,".** ",x,"</div>")
    x = paste0("**Table ",ifelse(supplementary,"S",""),capTabNo,".** ",x,"")
    if( id ) x <- paste0(x, "<a id='Tab",ifelse(supplementary,"S",""),capTabNo,"'/>");
    capTabNo <<- capTabNo + 1
  }; x
}
#Function to add the Figure Number
capFig = function(x, show_R_version=TRUE, show_package_versions=NULL, is_map=FALSE, supplementary=FALSE, id=TRUE)
{
  if(outputFormat == 'html')
  {
    x <- paste0("**Figure ",ifelse(supplementary,"S",""),capFigNo,".** ",x,"");
    if( show_R_version || (!is.null(show_package_versions) && length(show_package_versions) > 0) )
    {
      x <- paste0(x, " Figure generated using ");
      if( show_R_version ) x <- paste0(x, stringr::str_replace(R.version.string, stringr::fixed("R "), "[`R`](https://www.r-project.org/) "));
      if( !is.null(show_package_versions) && length(show_package_versions) > 0 )
      {
        x <- paste0(x, ifelse( show_R_version, " and ", " "));
        x <- paste0(x, ifelse( length(show_package_versions) > 1, "packages ", "package "));
        x <- paste0(x, paste0(vapply(show_package_versions, function(x) paste0("`",x,"`", " (version ", packageVersion(x),")"), character(1)), collapse=", "), ".");
      }
      if( is_map ) x <- paste0(x, " Maps are using public domain data from the [Natural Earth project](https://www.naturalearthdata.com/) as provided by the `R` package `maps`.");
      if( id ) x <- paste0(x, "<a id='Fig",ifelse(supplementary,"S",""),capFigNo,"'/>");
    }
    capFigNo <<- capFigNo + 1;
  }; 
  x;
}

```

```{r}
# Store the labels that we want to use for the different categories
categoryLabels <- c('Visual Artists', 'Non Artists', 'Human Inspired GenAI', 'Self-Guided GenAI')

scaleVariables <- c('liking', 'vividness', 'originality', 'aesthetics', 'curiosity',  'creativity')
labels <- c('Liking', 'Vividness', 'Originality', 'Aesthetics', 'Curiosity', 'Creativity')


selected_colours <- viridis(6)[c(1:3, 5)]

coloursRaters <- plasma(6)[c(1,3,4)]

```

```{r}
eval_data_cleaning <- FALSE
```


# Loading and preparing the data

The data for this study comes from different sources. First, we have the ratings provided by humam participants, which contain also their demographic information. To this information I will add more detailed data regarding their AReA score. Second, we have the ratings obtained from GPT-4O, both with and without human guidance. Finally, we also performed a "Turing test" where new participants were asked to identify the origin (human or GenAI) of the stimuli used in the rating task.

```{r load human ratings, eval = eval_data_cleaning}

bd <- read_excel('data/raw/combined_data_edited.xlsx') # This excel file was already pre-processed by Silvia

# Change names to something shorter 
names(bd) <- c('participantId', 'age', 'gender', 'yearsEducation', 'imageId', 
               'category', 'areaScore', 'liking', 'vividness', 'originality', 
               'aesthetics', 'curiosity', 'creativityScore')



bd$category <- factor(bd$category, labels=categoryLabels) 

# We generate a new category variable that divides the sample between Humans and GenAI without considering the subgroups.
bd$category2 <- factor(ifelse(bd$category %in% categoryLabels[1:2], 'Humans', 'GenAI'), levels=c('Humans', 'GenAI'))


## Cleaning demographic variables

# Let's take a look at the gender of participants:
#table(bd$gender) 

# We see that the contents of the variable are all over the place: participants 
# have used different ways to input their gender. We must change them in order to sort
# all results into four variables: male, female, non-binary and other 

bd$gender <- tolower(bd$gender) # switch all to lower case to remove categories

bd$gender[grepl('f', bd$gender)] <- 'Female' # All results that contain 'f' are female
bd$gender[grepl('mujer', bd$gender)] <- 'Female'

bd$gender[grepl('h', bd$gender)] <- 'Male' # All results that contain 'h' are male
bd$gender[startsWith(bd$gender, 'm')] <- 'Male' # Now, all results that start with 'm' are male (we have already classified 'Mujer')

bd$gender[bd$gender=='otro'] <- 'Other'
bd$gender[bd$gender=='non binary'] <- 'Non-binary'

bd$gender <- factor(bd$gender) 

# Fix some typos
bd$age[bd$age > 100] <- NA
bd$yearsEducation[bd$yearsEducation > 100] <- NA




# Generate separate database with only demographic data. This will make describing our sample easier
dm <- unique(bd[,c('participantId', 'age', 'gender', 'yearsEducation', 'areaScore')])




# Although Silvia's data file contains an overall AREA score, we want to have the individual 
# questions in case we want to make extra analysis with this information. 

# The raw data to calculate the AREA score is in the original individual files. Here we load them and extract the
# information regarding the individual questions from the AREA questionnaire. This will allow us to 
# analyze these data in depth and re-calculate the final score using PCA.

filesRaw <- list.files(path='data/raw/XLSX TOTAL UPDATED/', pattern='*.xlsx')

bdArea <- c()
missingIds <- c()

for (file in filesRaw) {
  
  xx <- read_excel(paste0('data/raw/XLSX TOTAL UPDATED/', file))
  if (sum(is.na(xx$PROLIFIC_PID)) > 0) {
    xx$PROLIFIC_PID <- unique(xx$PROLIFIC_PID)[1]
  }
  xx <- na.omit(xx[,c('PROLIFIC_PID', 'AreaSlider.response')])
  names(xx) <- c('participantId', 'areaAnswer')
  xx$areaAnswer <- as.numeric(substr(xx$areaAnswer, 1, 1))
  xx$areaQuestion <- paste0('q', formatC(1:14, width = 2, flag = "0"))
  xx <- spread(xx, areaQuestion, areaAnswer)
  
  
  bdArea <- rbind(bdArea, xx)
}
 rm(filesRaw, xx)

bdArea$appreciation <- (bdArea$q01 + bdArea$q02 + bdArea$q03 + 
                          bdArea$q04 + bdArea$q06 + bdArea$q09 + 
                          bdArea$q13 + bdArea$q14)/8

bdArea$experience <- (bdArea$q08 + bdArea$q11 + bdArea$q12 + bdArea$q13)/4

bdArea$behaviour <- (bdArea$q05 + bdArea$q07 + bdArea$q10)/3

dm <- merge(dm, bdArea, by='participantId', all=T)

# We will also generate a new variable dividing the sample into high and low area scores according to the sample median
dm$areaGroup <- factor(ifelse(dm$areaScore <= median(dm$areaScore), 'Low', 'High'))
bd$areaGroup <- factor(ifelse(bd$areaScore <= median(dm$areaScore), 'Low', 'High'))


save(dm, file ='data/processed/demographics.RData')
rm(bdArea)

```



```{r load gpt ratings, eval = eval_data_cleaning}


## Load data ####
# The results are stored in separate .xlsx files, one for each image, containing 30 ratings per image
# The code below gathers all these files and generates a single dataframe


filesRaw <- list.files(path='data/raw/results-gpt/', pattern='*.xlsx')

bdGpt <- c()

for (file in filesRaw) {
  
  xx <- read_excel(paste0('data/raw/results-gpt/', file))
  xx$id <- 1:nrow(xx)
  
  bdGpt <- rbind(bdGpt, xx)
}

names(bdGpt) <- c('category', 'imageId', 'liking', 'vividness', 'originality', 'aesthetics', 'curiosity', 'gptId')
rm(filesRaw, xx)

## Factor category ####
bdGpt$category <- factor(bdGpt$category, labels=categoryLabels) # CHECK WITH SILVIA: ADD LABELS
bdGpt$category2 <- factor(ifelse(bdGpt$category %in% categoryLabels[1:2], 'Humans', 'GenAI'), levels=c('Humans', 'GenAI'))






```



```{r guided GPT ratings, eval = eval_data_cleaning}

# Following one of the reviewer's comments we decided to run a new batch of GPT ratings, this time, using 
# some of the human ratings as guidance. 

# In this case, the data are stored in separate .xlsx files. 
# Some of the files were empty due to a problem with the GPT script. In this first loop I 
# identify which files are empty and remove them from the files list that will be imported later.

filesRaw <- list.files(path='data/raw/results_with_refimages/results_with_refimages/', pattern='*.xlsx', recursive=TRUE)

bdGptRef <- c()
listResponses <- c()

for (file in filesRaw) {
  
  xx <- read_excel(paste0('data/raw/results_with_refimages/results_with_refimages/', file))

  listResponses <- rbind(listResponses, c(file, nrow(xx)))
  
}


listResponsesDb <- as.data.frame(listResponses)
listResponsesDb$V2 <- as.numeric(listResponsesDb$V2)

to_remove <- listResponsesDb$V1[listResponsesDb$V2 == 0]

filesRawComplete <- filesRaw[!(filesRaw %in% to_remove)] # keep only files with responses in them

# Note: not all of the files will have 30 ratings in them. Due to problems with the GPT script,
# some of the images have raings divided into different files. This is not a problem since in the 
# final dataset they will be identified by image name.

bdGptRef <- c()


for (file in filesRawComplete) {
  
  xx <- read_excel(paste0('data/raw/results_with_refimages/results_with_refimages/', file))
  xx$id <- 1:nrow(xx)
  
  bdGptRef <- rbind(bdGptRef, xx)
  

}

names(bdGptRef)[1:7] <- c('category', 'imageId', 'liking', 'vividness', 'originality', 'aesthetics', 'curiosity')

bdGptRef$category <- ifelse(bdGptRef$category=='category1', categoryLabels[1], 
                            ifelse(bdGptRef$category=='category2', categoryLabels[2], 
                                   ifelse(bdGptRef$category=='category3', categoryLabels[3], categoryLabels[4])))

bdGptRef$category <- factor(bdGptRef$category, levels = categoryLabels)


bdGptRefWhole <- bdGptRef

bdGptRef <- bdGptRef[,1:7]

bdGptRef$rater <- 'GPT-REF'

```



```{r merge all databases, eval = eval_data_cleaning}

# One of the Visual artists that generated the images was missclassified as a Non-artist. Here I change
# the category of their pictures

imagesToChange <- c('P29S1.png', 'P29S2.png','P29S3.png','P29S4.png','P29S5.png',
                    'P29S6.png','P29S7.png','P29S8.png','P29S9.png','P29S10.png','P29S11.png')

bd$category[bd$imageId %in% imagesToChange] <- 'Visual Artists'
bdGpt$category[bdGpt$imageId %in% imagesToChange] <- 'Visual Artists'
bdGptRef$category[bdGptRef$imageId %in% imagesToChange] <- 'Visual Artists'

rm(imagesToChange)



bdHuman <- bd[c('imageId', 'category', 'liking', 'vividness', 'originality', 'aesthetics', 'curiosity', 'participantId')]
bdGpt <- bdGpt[c('imageId', 'category', 'liking', 'vividness', 'originality', 'aesthetics', 'curiosity')]
bdGuidedGpt <- bdGptRef[c('imageId', 'category', 'liking', 'vividness', 'originality', 'aesthetics', 'curiosity')]

bdHuman$raterType <- 'Human'
bdGpt$raterType <- 'GPT'
bdGuidedGpt$raterType <- 'GuidedGPT'

bdGpt$participantId <- 'GPT'
bdGuidedGpt$participantId <- 'GuidedGPT'


bd <- rbind(bdHuman, bdGpt,bdGuidedGpt)

bd$category2 <- factor(ifelse(bd$category %in% categoryLabels[1:2], 'Humans', 'GenAI'), levels=c('Humans', 'GenAI'))
bd$raterType <- factor(bd$raterType, levels=c('Human', 'GPT', 'GuidedGPT'))

```


```{r generate creativity score, eval = eval_data_cleaning}

# Modified from Dan

  crea.1f <- "
    creativity =~ liking + vividness + originality + aesthetics + curiosity;
  ";
  
  # configurational invariance:
  crea.1f.fit <- cfa(crea.1f, data=bd, group="raterType", se="bootstrap", test="bootstrap");
  # weak invariance:
  crea.1f.fit.wi <- cfa(crea.1f, data=bd, group="raterType", group.equal="loadings", se="bootstrap", test="bootstrap");
  # strong invariance:
  crea.1f.fit.si <- cfa(crea.1f, data=bd, group="raterType", group.equal=c("intercepts", "loadings"), se="bootstrap", test="bootstrap");
  # model comparison tests
  lavTestLRT(crea.1f.fit, crea.1f.fit.wi, crea.1f.fit.si); # both p-values are signif. ->  factor loadings and latent means are different between judge types! 
  
  # pairwise comparisons between types of judges:
  # H vs AI:
  d <- bd[bd$raterType %in% c("Human","GPT"),] %>% mutate(raterType=as.character(raterType));
  crea.1f.ha.fit <- cfa(crea.1f, data=d, group="raterType", se="bootstrap", test="bootstrap");
  crea.1f.ha.fit.wi <- cfa(crea.1f, data=d, group="raterType", group.equal="loadings", se="bootstrap", test="bootstrap");
  crea.1f.ha.fit.si <- cfa(crea.1f, data=d, group="raterType", group.equal=c("intercepts", "loadings"), se="bootstrap", test="bootstrap");
  lavTestLRT(crea.1f.ha.fit, crea.1f.ha.fit.wi, crea.1f.ha.fit.si); # both p-values are signif. ->  factor loadings and latent means are different between judge types! 
  
  # H vs gAI:
  d <- bd[bd$raterType %in% c("Human","GuidedGPT"),] %>% mutate(raterType=as.character(raterType));
  crea.1f.hg.fit <- cfa(crea.1f, data=d, group="raterType", se="bootstrap", test="bootstrap");
  crea.1f.hg.fit.wi <- cfa(crea.1f, data=d, group="raterType", group.equal="loadings", se="bootstrap", test="bootstrap");
  crea.1f.hg.fit.si <- cfa(crea.1f, data=d, group="raterType", group.equal=c("intercepts", "loadings"), se="bootstrap", test="bootstrap");
  lavTestLRT(crea.1f.hg.fit, crea.1f.hg.fit.wi, crea.1f.hg.fit.si); # both p-values are signif. ->  factor loadings and latent means are different between judge types! 
  
  # AI vs gAI:
  d <- bd[bd$raterType %in% c("GPT","GuidedGPT"),] %>% mutate(raterType=as.character(raterType));
  crea.1f.ag.fit <- cfa(crea.1f, data=d, group="raterType", se="bootstrap", test="bootstrap");
  crea.1f.ag.fit.wi <- cfa(crea.1f, data=d, group="raterType", group.equal="loadings", se="bootstrap", test="bootstrap");
  crea.1f.ag.fit.si <- cfa(crea.1f, data=d, group="raterType", group.equal=c("intercepts", "loadings"), se="bootstrap", test="bootstrap");
  lavTestLRT(crea.1f.ag.fit, crea.1f.ag.fit.wi, crea.1f.ag.fit.si); # both p-values are signif. ->  factor loadings and latent means are different between judge types! 
  
  # separately for types of judges:
  crea.1f.fit.hum <- cfa(crea.1f, data=bd[bd$raterType=="Human",], se="bootstrap", test="bootstrap");

  crea.1f.fit.ai <- cfa(crea.1f, data=bd[bd$raterType=="GPT",], se="bootstrap", test="bootstrap");

  crea.1f.fit.gai <- cfa(crea.1f, data=bd[bd$raterType=="GuidedGPT",], se="bootstrap", test="bootstrap");

  # collect and save these results:
  crea1FA <- list("model"=crea.1f, 
                  "fit"=crea.1f.fit, "invariance.tests"=lavTestLRT(crea.1f.fit, crea.1f.fit.wi, crea.1f.fit.si), 
                  "fit.h.ai"  =crea.1f.ha.fit, "invariance.tests.h.ai"=lavTestLRT(crea.1f.ha.fit, crea.1f.ha.fit.wi, crea.1f.ha.fit.si), 
                  "fit.h.gai" =crea.1f.hg.fit, "invariance.tests.h.gai"=lavTestLRT(crea.1f.hg.fit, crea.1f.hg.fit.wi, crea.1f.hg.fit.si), 
                  "fit.ai.gai"=crea.1f.ag.fit, "invariance.tests.ai.gai"=lavTestLRT(crea.1f.ag.fit, crea.1f.ag.fit.wi, crea.1f.ag.fit.si), 
                  "fit.hum"=crea.1f.fit.hum, "fit.ai"=crea.1f.fit.ai, "fit.gai"=crea.1f.fit.gai);



crea1FA.pred <- lavPredict(crea1FA$fit, type="lv", method="regression", transform=TRUE, append.data=TRUE, assemble=TRUE);
# check that the order is correct:
all(vapply(1:nrow(bd), function(i) bd$raterType[i] == crea1FA.pred$raterType[i] && bd$liking[i] == crea1FA.pred$liking[i] && 
             bd$vividness[i] == crea1FA.pred$vividness[i] && bd$originality[i] == crea1FA.pred$originality[i] && bd$aesthetics[i] == crea1FA.pred$aesthetics[i] && 
             bd$curiosity[i] == crea1FA.pred$curiosity[i], logical(1)));
bd$creativity <- crea1FA.pred$creativity;
bd$creativity <- 1 + (7-1)*(bd$creativity - min(bd$creativity, na.rm=TRUE))/(max(bd$creativity, na.rm=TRUE) - min(bd$creativity, na.rm=TRUE)) 
# convert it to a 1 to 7 scale to make it directly comparable with the original scales



save(bd, file = 'data/processed/database-whole.RData')
```


```{r turing test data, eval = eval_data_cleaning}

# load list of images used for the turing test

turingTestImages <- read.delim("data/raw/randomised_image_list.txt", header = F)

turingTestImages$images <- sapply(strsplit(turingTestImages$V1, split = "/"), `[`, 7)
turingTestImages$category <- sapply(strsplit(turingTestImages$V1, split = "/"), `[`, 6)

turingTestImages$item_number <- 1:nrow(turingTestImages)

# load turing test results

turing <- read_excel("data/raw/TOTAL TURING TEST RESULTS.xlsx" )

# create vector of true responses
responses <- cbind(1:48, 
                   c("A. Human", "B. AI", "A. Human", "A. Human", "A. Human", "B. AI", "B. AI", "B. AI", "B. AI", "A. Human", "B. AI", "B. AI", "A. Human", "B. AI", "A. Human", "B. AI", "A. Human", "A. Human", "A. Human", "B. AI", "B. AI", "A. Human", "A. Human", "B. AI", "A. Human", "A. Human", "B. AI", "B. AI", "A. Human", "A. Human", "B. AI", "B. AI", "A. Human", "B. AI", "A. Human", "B. AI", "B. AI", "A. Human", "B. AI", "A. Human", "A. Human", "A. Human", "B. AI", "A. Human", "B. AI", "B. AI", "A. Human", "B. AI"))

names(turing) <- c('participantId', 'studentId', 'score_pc', 'score_n', 'gender', 'age', paste0(rep(c('item_', 'confidence_'), times=48), rep(1:48, each=2)), paste0(rep("q_", times=13), 1:13))



turing <- turing %>%
  pivot_longer(
    cols = matches("^(item|confidence)_"),
    names_to = c(".value", "item_number"),
    names_pattern = "(item|confidence)_(\\d+)"
  )


responses <- data.frame(responses)
names(responses) <- c('item_number', 'answer')

turing <- merge(turing, responses, by='item_number', all.x=T)
turing$correct <- factor(ifelse(turing$item == turing$answer, 1, 0), levels=c(0, 1), labels=c('Incorrect', 'Correct'))
turing$confidence <- as.numeric(substr(turing$confidence, 4, 4))

turing <- merge(turing, turingTestImages, by='item_number', all.x=T)

save(turing, file = 'data/processed/turing-test-data.RData')

rm(turingTestImages)

```


```{r load already processed data, eval = !eval_data_cleaning}

load('data/processed/database-whole.RData')
load('data/processed/demographics.RData')
load('data/processed/turing-test-data.RData')

```



```{r histogram figure s1}
bd.long <- bd %>% melt(measure.vars=c("liking", "vividness", "originality", "aesthetics", "curiosity"), variable.name="question", value.name="value");

bd.long$question <- factor(bd.long$question, levels=c("liking", "vividness", "originality", "aesthetics", "curiosity"), labels=c("Liking", "Vividness", "Originality", "Aesthetics", "Curiosity"))


bd.long$raterType <- factor(bd.long$raterType, levels=c('Human', 'GPT', 'GuidedGPT'), labels=c('Human', 'GPT', 'Guided GPT'))
ggplot(bd.long, aes(x=value)) + 
               geom_histogram(aes(fill=raterType)) + theme_classic() + 
               facet_grid(raterType ~ question) + theme(legend.position="none", text=element_text(family='serif')) + xlab('Value') + ylab('Count') + scale_fill_manual(values=coloursRaters)
```




```{r figura s2}

# Aquesta figura fa servir els pesos etc del PCA

d.gai <- bd[bd$raterType=="GuidedGPT",c("liking", "vividness", "originality", "aesthetics", "curiosity")];
d.hum <- bd[bd$raterType=="Human",c("liking", "vividness", "originality", "aesthetics", "curiosity")];
d.ai <- bd[bd$raterType=="GPT",c("liking", "vividness", "originality", "aesthetics", "curiosity")];

names(d.gai) <- names(d.hum) <- names(d.ai) <- c('Liking', 'Vividness', 'Originality', 'Aesthetics', 'Curiosity')

t_pca.gai <- prcomp(d.gai, scale.=TRUE, center=TRUE);
t_pca.ai <- prcomp(d.ai, scale.=TRUE, center=TRUE);
t_pca.hum <- prcomp(d.hum, scale.=TRUE, center=TRUE);


eig <- get_eig(t_pca.hum)

# Step 3: Build the plot manually
eig_df <- data.frame(
  dim = factor(paste0("Dim ", seq_along(eig$variance.percent)), 
               levels = paste0("Dim ", seq_along(eig$variance.percent))),
  percent = eig$variance.percent
)

(plotPcaCompHuman <- ggplot(eig_df, aes(x = dim, y = percent)) +
    geom_bar(stat = "identity", 
             fill = coloursRaters[1], 
             color = coloursRaters[1]) +
    geom_text(aes(label = paste0(round(percent, 1), "%")),
              vjust = -0.5,
              size = 3,               # <== change label font size here
              family = "serif") +
    theme_minimal(base_family = "serif") +
    theme(
      axis.text = element_text(size = 10),
      axis.title = element_text(size = 12),
      plot.title = element_text(size = 12, face = "bold")
    ) +
    labs(
      x = "Dimensions",
      y = "Percentage of explained variance",
      title = "Humans"
    ))



eig <- get_eig(t_pca.gai)

# Step 3: Build the plot manually
eig_df <- data.frame(
  dim = factor(paste0("Dim ", seq_along(eig$variance.percent)), 
               levels = paste0("Dim ", seq_along(eig$variance.percent))),
  percent = eig$variance.percent
)

(plotPcaCompGuidedAI <- ggplot(eig_df, aes(x = dim, y = percent)) +
    geom_bar(stat = "identity", 
             fill = coloursRaters[3], 
             color = coloursRaters[3]) +
    geom_text(aes(label = paste0(round(percent, 1), "%")),
              vjust = -0.5,
              size = 3,               # <== change label font size here
              family = "serif") +
    theme_minimal(base_family = "serif") +
    theme(
      axis.text = element_text(size = 10),
      axis.title = element_text(size = 12),
      plot.title = element_text(size = 12, face = "bold")
    ) +
    labs(
      x = "Dimensions",
      y = "Percentage of explained variance",
      title = "Guided GPT"
    ))



eig <- get_eig(t_pca.ai)

# Step 3: Build the plot manually
eig_df <- data.frame(
  dim = factor(paste0("Dim ", seq_along(eig$variance.percent)), 
               levels = paste0("Dim ", seq_along(eig$variance.percent))),
  percent = eig$variance.percent
)

(plotPcaCompAI <- ggplot(eig_df, aes(x = dim, y = percent)) +
    geom_bar(stat = "identity", 
             fill = coloursRaters[2], 
             color = coloursRaters[2]) +
    geom_text(aes(label = paste0(round(percent, 1), "%")),
              vjust = -0.5,
              size = 3,               # <== change label font size here
              family = "serif") +
    theme_minimal(base_family = "serif") +
    theme(
      axis.text = element_text(size = 10),
      axis.title = element_text(size = 12),
      plot.title = element_text(size = 12, face = "bold")
    ) +
    labs(
      x = "Dimensions",
      y = "Percentage of explained variance",
      title = "GPT"
    ))






pca <- t_pca.hum

eig <- summary(pca)$importance[2, ] * 100

var <- get_pca_var(pca)

df <- data.frame(
  x = var$coord[, 1],
  y = var$coord[, 2],
  name = rownames(var$coord)
)

(plotWeightsHuman <-ggplot(df, aes(x, y, label = name)) +
  geom_segment(
    aes(x = 0, y = 0, xend = x, yend = y),
    arrow = arrow(length = unit(0.2, "cm")),
    color = "black"
  ) +
  geom_text_repel(size = 4, family = "serif") +
  labs(
    x = paste0("Dim 1 (", round(eig[1], 1), "%)"),
    y = paste0("Dim 2 (", round(eig[2], 1), "%)")
  ) +
  theme_minimal(base_family = "serif")+ 
    theme(panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.border = element_blank(),
          panel.background = element_blank()) + 
    geom_hline(yintercept=0, linetype="dashed", color = "black") + 
    geom_vline(xintercept=0, linetype="dashed", color = "black")  )




pca <- t_pca.gai

eig <- summary(pca)$importance[2, ] * 100

var <- get_pca_var(pca)

df <- data.frame(
  x = var$coord[, 1],
  y = var$coord[, 2],
  name = rownames(var$coord)
)

(plotWeightsGuidedAI <-ggplot(df, aes(x, y, label = name)) +
  geom_segment(
    aes(x = 0, y = 0, xend = x, yend = y),
    arrow = arrow(length = unit(0.2, "cm")),
    color = "black"
  ) +
  geom_text_repel(size = 4, family = "serif") +
  labs(
    x = paste0("Dim 1 (", round(eig[1], 1), "%)"),
    y = paste0("Dim 2 (", round(eig[2], 1), "%)")
  ) +
  theme_minimal(base_family = "serif")+ 
    theme(panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.border = element_blank(),
          panel.background = element_blank()) + 
    geom_hline(yintercept=0, linetype="dashed", color = "black") + 
    geom_vline(xintercept=0, linetype="dashed", color = "black")  )


pca <- t_pca.ai

eig <- summary(pca)$importance[2, ] * 100

var <- get_pca_var(pca)

df <- data.frame(
  x = var$coord[, 1],
  y = var$coord[, 2],
  name = rownames(var$coord)
)

(plotWeightsAI <-ggplot(df, aes(x, y, label = name)) +
  geom_segment(
    aes(x = 0, y = 0, xend = x, yend = y),
    arrow = arrow(length = unit(0.2, "cm")),
    color = "black"
  ) +
  geom_text_repel(size = 4, family = "serif") +
  labs(
    x = paste0("Dim 1 (", round(eig[1], 1), "%)"),
    y = paste0("Dim 2 (", round(eig[2], 1), "%)")
  ) +
  theme_minimal(base_family = "serif") + 
    theme(panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.border = element_blank(),
          panel.background = element_blank()) + 
    geom_hline(yintercept=0, linetype="dashed", color = "black") + 
    geom_vline(xintercept=0, linetype="dashed", color = "black")  )

ggarrange(
ggarrange(plotPcaCompHuman, plotWeightsHuman, labels=c('A', 'B')),
ggarrange(plotPcaCompAI, plotWeightsAI, labels=c('C', 'D')),
ggarrange(plotPcaCompGuidedAI, plotWeightsGuidedAI, labels=c('E', 'F')),
nrow=3)

```


# 0- Response to the reviewer


## Comment 5


Original comment: **It would strengthen the study to include multiple human raters evaluating the same images, enabling assessment of inter-rater reliability and reproducibility.**

```{r}
humans <- subset(bd, bd$raterType=='Human')

```


```{r calculate distances and cohens kappas}

# # Removing the names of participants and images and replacing with numbers will make the output more legible
# 
# humans$participantId <- as.numeric(factor(humans$participantId))
# humans$imageId <- as.numeric(factor(humans$imageId))
# 
# humans <- humans[,c('participantId', 'imageId', 'liking', 'vividness', 'originality', 'aesthetics', 'curiosity', 'creativity')]
# 
# # Generate empty matrices to store the results of the distances between ratings, Cohen's and AREA scores
# 
# resultsDistances <- lapply(1:6, matrix, data= NA, nrow=length(unique(humans$participantId)), ncol=length(unique(humans$participantId)))
# 
# resultsCohen <- lapply(1:6, matrix, data= NA, nrow=length(unique(humans$participantId)), ncol=length(unique(humans$participantId)))
# 
# commonPictures <- matrix(NA,  nrow=length(unique(humans$participantId)), ncol=length(unique(humans$participantId)))
# 
# areaDistances <- matrix(NA,  nrow=length(unique(humans$participantId)), ncol=length(unique(humans$participantId)))
# 
# # Loop between all pairs of participants
# for (i in 1:(length(unique(humans$participantId))-1)) {
#   for (s in (i+1):255-i) {
#     j <- i + s
# 
#     # Keep only the data from the two selected participants
#     subset <- subset(humans, humans$participantId %in% c(i, j))
#     # Keep only the images that have been rated by both participants
#     subsetClean <- na.omit(dcast(melt(subset, id.vars=c("imageId", "participantId")), imageId~variable+participantId))
# 
#     if (nrow(subsetClean) > 0) { #if the participants have at least one image in common, calculate distances and Cohen's kappa values
# 
#       # To generate the distences, we use the mean of the absolute differences between all ratings for each pair of participants
#       resultsDistances[[1]][i, j] <- mean(abs(subsetClean[,2] - subsetClean[,3]))
#       resultsDistances[[2]][i, j] <- mean(abs(subsetClean[,4] - subsetClean[,5]))
#       resultsDistances[[3]][i, j] <- mean(abs(subsetClean[,6] - subsetClean[,7]))
#       resultsDistances[[4]][i, j] <- mean(abs(subsetClean[,8] - subsetClean[,9]))
#       resultsDistances[[5]][i, j] <- mean(abs(subsetClean[,10] - subsetClean[,11]))
#       resultsDistances[[6]][i, j] <- mean(abs(subsetClean[,12] - subsetClean[,13]))
# 
#       # We use the weighted version of Cohen's data, since we are dealing with ordinal variables (except the Creativity score, which is continuous)
#       resultsCohen[[1]][i, j] <- kappa2(subsetClean[,c(2,3)])$value
#       resultsCohen[[2]][i, j] <- kappa2(subsetClean[,c(4,5)])$value
#       resultsCohen[[3]][i, j] <- kappa2(subsetClean[,c(6,7)])$value
#       resultsCohen[[4]][i, j] <- kappa2(subsetClean[,c(8,9)])$value
#       resultsCohen[[5]][i, j] <- kappa2(subsetClean[,c(10,11)])$value
#       resultsCohen[[6]][i, j] <- kappa(subsetClean[,c(12,13)])
# 
# 
#     }
# 
#     # The number of common pictures is equal to the number of rows in the subsetted data, since we have removed missing values
#     commonPictures[i, j] <- nrow(subsetClean)
# 
# 
#     areaDistances[i, j] <- dm$areaScore[i] - dm$areaScore[j]
# 
#   }
# 
# }
# 
# 
# 
# save(resultsDistances, file = 'output/results-distances.RData')
# save(resultsCohen, file = 'output/results-cohen.RData')
# save(areaDistances, file = 'output/area-distances.RData')
# save(commonPictures, file = 'output/common-pictures.RData')
# 



load(file = 'output/results-distances.RData')
load(file = 'output/results-cohen.RData')
load(file = 'output/area-distances.RData')
load(file = 'output/common-pictures.RData')

```

We will provide some measures to assess Inter-rater reliability and reproducibility. 

The main problem is that most of the participants in the sample received completely different image datasets to rate, making it difficult to calculate inter-rater measures.

Here we can see the number of images in common for each pair of participants:

```{r raw histogram common pictures}
# We generate a dataframe because we need it for ggplot

dfCommonPictures <- as.data.frame(as.vector(commonPictures)[!is.na(as.vector(commonPictures))])
names(dfCommonPictures) <- 'value'

ggplot(dfCommonPictures, aes(x=value)) + 
  geom_histogram() +  theme_minimal() +
  labs(x='Number of common pictures', y='Number of pairs')
```


We see that the distribution shows some instances where participants rated the same 120 images. We plot again, removing those participants:

```{r selected histogram common pictures}

dfCommonPictures100 <- subset(dfCommonPictures, dfCommonPictures$value < 100)

limitsCommonPictures100 <- c(mean(dfCommonPictures100$value, na.rm=T) - sd(dfCommonPictures100$value, na.rm=T),
                             mean(dfCommonPictures100$value, na.rm=T) + sd(dfCommonPictures100$value, na.rm=T))

ggplot(dfCommonPictures100, aes(x=value)) + 
  geom_histogram(bins=40) +  theme_minimal() +
  labs(x='Number of common pictures', y='Number of pairs')+
  geom_vline(xintercept = mean(dfCommonPictures100$value, na.rm=T), colour='red')+
  geom_vline(xintercept = limitsCommonPictures100[1], colour='blue')+
  geom_vline(xintercept = limitsCommonPictures100[2], colour='blue')
 
```

We can calculate, for the whole sample, Krippendorff's alpha:

```{r kripp.alpha loop}
# # For the kripp.alpha() function we need a matrix with where each column is a rater and each row is an item:
# 
# matrices <- list()
# irrResults <- c()
# variables <- c('liking', 'vividness', 'originality', 'aesthetics', 'curiosity')
# 
# for (v in 1:length(variables)) {
# 
#   matrices[[v]] <- humans[,c('participantId', 'imageId', variables[v])]
#   matrices[[v]] <- spread(matrices[[v]], imageId, variables[v])
# 
#   krippOutput <-  kripp.alpha(t(t(matrices[[v]][,-1])), method = "ordinal") # We need to specify that these are ordinal variables
#   irrResults <- rbind(irrResults, c(variables[v], krippOutput$subjects, krippOutput$raters, krippOutput$value))
# 
# }
# 
# 
# rownames(irrResults) <- irrResults[,1]
# irrResults <- irrResults[,-1]
# colnames(irrResults) <- c('Number of images', 'Number of raters', 'Alpha')
# 
# save(irrResults, file = 'output/irr-results.RData') # This step is quite time-consuming, so I save the output and just pull it. If we change anything from the code above we will need to run this again and re-store the new version

load(file = 'output/irr-results.RData')

kable(irrResults)

```

Since we have so missing values in the matrix, however, the results obtained are unreliable. We can also computate, for each pair of participants, their Cohen's d value. However, this measure will also be quite unreliable since so many pairs have very few images in common.

```{r histograms cohen}


dataframeCohen <- as.data.frame(rbind(cbind(as.vector(resultsCohen[[1]])[!is.na(as.vector(resultsCohen[[1]]))], 'liking'),
                                          cbind(as.vector(resultsCohen[[2]])[!is.na(as.vector(resultsCohen[[2]]))], 'vividness'),
                                          cbind(as.vector(resultsCohen[[3]])[!is.na(as.vector(resultsCohen[[3]]))], 'originality'),
                                          cbind(as.vector(resultsCohen[[4]])[!is.na(as.vector(resultsCohen[[4]]))], 'aesthetics'),
                                          cbind(as.vector(resultsCohen[[5]])[!is.na(as.vector(resultsCohen[[5]]))], 'curiosity')))


names(dataframeCohen) <- c('cohen', 'category')
dataframeCohen$cohen <- as.numeric(dataframeCohen$cohen)
dataframeCohen$category <- factor(dataframeCohen$category, levels=c('liking', 'vividness', 'originality', 'aesthetics', 'curiosity'))


ggplot(dataframeCohen, aes(x=cohen)) + 
  geom_histogram() + facet_grid(~category) + theme_classic()

```

A more reliable way of assessing inter-rater reliability in a case like ours, with unequal raters per item and missingness, is using a mixed-effects model to perform variance decomposition. This allows us to estimate the percentage of variance in the ratings attributed to the effect of the raters and the images separately. 

### Mixed effects model

```{r mixed effects model for inter rater reliability}

model_liking <- lmer(liking ~ 1 + (1|imageId) + (1|participantId), data=humans)
vc_liking <- as.data.frame(VarCorr(model_liking))
sigma_image_liking <- vc_liking[vc_liking$grp=="imageId","vcov"]
sigma_rater_liking <- vc_liking[vc_liking$grp=="participantId","vcov"]
sigma_err_liking   <- attr(VarCorr(model_liking),"sc")^2
reliability_liking <- sigma_image_liking /
  (sigma_image_liking + sigma_rater_liking + sigma_err_liking)

model_vividness <- lmer(vividness ~ 1 + (1|imageId) + (1|participantId), data=humans)
vc_vividness <- as.data.frame(VarCorr(model_vividness))
sigma_image_vividness <- vc_vividness[vc_vividness$grp=="imageId","vcov"]
sigma_rater_vividness <- vc_vividness[vc_vividness$grp=="participantId","vcov"]
sigma_err_vividness   <- attr(VarCorr(model_vividness),"sc")^2
reliability_vividness <- sigma_image_vividness /
  (sigma_image_vividness + sigma_rater_vividness + sigma_err_vividness)

model_originality <- lmer(originality ~ 1 + (1|imageId) + (1|participantId), data=humans)
vc_originality <- as.data.frame(VarCorr(model_originality))
sigma_image_originality <- vc_originality[vc_originality$grp=="imageId","vcov"]
sigma_rater_originality <- vc_originality[vc_originality$grp=="participantId","vcov"]
sigma_err_originality   <- attr(VarCorr(model_originality),"sc")^2
reliability_originality <- sigma_image_originality /
  (sigma_image_originality + sigma_rater_originality + sigma_err_originality)

model_aesthetics <- lmer(aesthetics ~ 1 + (1|imageId) + (1|participantId), data=humans)
vc_aesthetics <- as.data.frame(VarCorr(model_aesthetics))
sigma_image_aesthetics <- vc_aesthetics[vc_aesthetics$grp=="imageId","vcov"]
sigma_rater_aesthetics <- vc_aesthetics[vc_aesthetics$grp=="participantId","vcov"]
sigma_err_aesthetics   <- attr(VarCorr(model_aesthetics),"sc")^2
reliability_aesthetics <- sigma_image_aesthetics /
  (sigma_image_aesthetics + sigma_rater_aesthetics + sigma_err_aesthetics)

model_curiosity <- lmer(curiosity ~ 1 + (1|imageId) + (1|participantId), data=humans)
vc_curiosity <- as.data.frame(VarCorr(model_curiosity))
sigma_image_curiosity <- vc_curiosity[vc_curiosity$grp=="imageId","vcov"]
sigma_rater_curiosity <- vc_curiosity[vc_curiosity$grp=="participantId","vcov"]
sigma_err_curiosity   <- attr(VarCorr(model_curiosity),"sc")^2
reliability_curiosity <- sigma_image_curiosity /
  (sigma_image_curiosity + sigma_rater_curiosity + sigma_err_curiosity)

print(cat('Liking: ', reliability_liking,  
             '\nVividness: ', reliability_vividness, 
             '\nOriginality: ', reliability_originality, 
             '\nAesthetics: ', reliability_aesthetics, 
             '\nCuriosity: ', reliability_curiosity, ''))

```

Reliability is very low (between 10 and 18), meaning that the proportion of total rating variance that is attributable to real differences between images is low. 

We can compute another measure of reliability using a split half technique, that will allow us to estimate how stable the image-level scores are. We will randomly split the ratings of each image into two halves and correlate the mean ratings of both subsets, repeating this operation in order to obtain a distribution and calculate a confidence interval. 

```{r split half function}

split_half_imagewise <- function(df,

                                 image_col = "imageId",

                                 rating_col = "rating",

                                 min_per_half = 2,

                                 n_repeats = 5000,

                                 method = c("spearman", "pearson"),

                                 seed = 42) {



  method <- match.arg(method)

  set.seed(seed)



  # Split df into a list of rating vectors by image

  ratings_by_image <- split(df[[rating_col]], df[[image_col]])



  corrs <- numeric(n_repeats)

  n_images_used <- integer(n_repeats)



  for (b in seq_len(n_repeats)) {



    halfA <- numeric(0)

    halfB <- numeric(0)



    for (img in names(ratings_by_image)) {

      x <- ratings_by_image[[img]]

      m <- length(x)



      # Need enough ratings to put at least min_per_half in each half

      if (m < 2 * min_per_half) next



      idx <- sample.int(m, m, replace = FALSE)

      a_idx <- idx[seq_len(floor(m / 2))]

      b_idx <- idx[(floor(m / 2) + 1):m]



      if (length(a_idx) < min_per_half || length(b_idx) < min_per_half) next



      halfA[img] <- mean(x[a_idx], na.rm = TRUE)

      halfB[img] <- mean(x[b_idx], na.rm = TRUE)

    }



    # Align images present in both halves (should be same keys, but be safe)

    common <- intersect(names(halfA), names(halfB))

    n_images_used[b] <- length(common)



    if (length(common) < 3) {

      corrs[b] <- NA_real_

    } else {

      corrs[b] <- suppressWarnings(

        cor(halfA[common], halfB[common], method = method, use = "complete.obs")

      )

    }

  }



  corrs <- corrs[!is.na(corrs)]

  n_images_used <- n_images_used[n_images_used > 0]



  # Spearman–Brown correction to estimate reliability of full aggregation

  sb <- (2 * corrs) / (1 + corrs)



  list(

    n_repeats_used = length(corrs),

    median_images_per_split = median(n_images_used),

    corr_median = median(corrs),

    corr_ci95 = as.numeric(quantile(corrs, c(0.025, 0.975), na.rm = TRUE)),

    spearman_brown_median = median(sb),

    spearman_brown_ci95 = as.numeric(quantile(sb, c(0.025, 0.975), na.rm = TRUE)),

    corrs = corrs,

    sb = sb

  )

} 

```

```{r split half generation and output}

# humans$rating <- humans$liking
# 
# result_liking <- split_half_imagewise(humans,
# 
# 
#                                min_per_half = 2,
# 
#                               n_repeats = 5000,
# 
#                                method = "spearman",
# 
#                                seed = 1)
# 
# # print(result_liking[c("n_repeats_used","median_images_per_split",
# #
# #                "corr_median","corr_ci95",
# #
# #                "spearman_brown_median","spearman_brown_ci95")])
# #
# 
# humans$rating <- humans$vividness
# 
# result_vividness <- split_half_imagewise(humans,
# 
# 
#                                min_per_half = 2,
# 
#                               n_repeats = 5000,
# 
#                                method = "spearman",
# 
#                                seed = 1)
# 
# # print(result_vividness[c("n_repeats_used","median_images_per_split",
# #
# #                "corr_median","corr_ci95",
# #
# #                "spearman_brown_median","spearman_brown_ci95")])
# #
# 
# humans$rating <- humans$originality
# 
# result_originality <- split_half_imagewise(humans,
# 
# 
#                                min_per_half = 2,
# 
#                               n_repeats = 5000,
# 
#                                method = "spearman",
# 
#                                seed = 1)
# #
# # print(result_originality[c("n_repeats_used","median_images_per_split",
# #
# #                "corr_median","corr_ci95",
# #
# #                "spearman_brown_median","spearman_brown_ci95")])
# 
# 
# humans$rating <- humans$aesthetics
# 
# result_aesthetics <- split_half_imagewise(humans,
# 
# 
#                                min_per_half = 2,
# 
#                               n_repeats = 5000,
# 
#                                method = "spearman",
# 
#                                seed = 1)
# 
# # print(result_aesthetics[c("n_repeats_used","median_images_per_split",
# #
# #                "corr_median","corr_ci95",
# #
# #                "spearman_brown_median","spearman_brown_ci95")])
# #
# 
# humans$rating <- humans$curiosity
# 
# result_curiosity <- split_half_imagewise(humans,
# 
# 
#                                min_per_half = 2,
# 
#                               n_repeats = 5000,
# 
#                                method = "spearman",
# 
#                                seed = 1)
# 
# # print(result_curiosity[c("n_repeats_used","median_images_per_split",
# #
# #                "corr_median","corr_ci95",
# #
# #                "spearman_brown_median","spearman_brown_ci95")])
# #
# 
# save(result_liking, file = 'output/split-half-liking.RData')
# save(result_vividness, file = 'output/split-half-vividness.RData')
# save(result_originality, file = 'output/split-half-originality.RData')
# save(result_aesthetics, file = 'output/split-half-aesthetics.RData')
# save(result_curiosity, file = 'output/split-half-curiosity.RData')


load('output/split-half-liking.RData')
load('output/split-half-vividness.RData')
load('output/split-half-originality.RData')
load('output/split-half-aesthetics.RData')
load('output/split-half-curiosity.RData')


```

```{r split half output}

rbind(
c('Liking', paste0(round(result_liking$corr_median,3), " (", round(result_liking$corr_ci95[1], 3), " - ", round(result_liking$corr_ci95[2], 3), ")"),
paste0(round(result_liking$spearman_brown_median,3), " (", round(result_liking$spearman_brown_ci95[1], 3), " - ", round(result_liking$spearman_brown_ci95[2], 3), ")")),
c('Vividness', paste0(round(result_vividness$corr_median,3), " (", round(result_vividness$corr_ci95[1], 3), " - ", round(result_vividness$corr_ci95[2], 3), ")"),
paste0(round(result_vividness$spearman_brown_median,3), " (", round(result_vividness$spearman_brown_ci95[1], 3), " - ", round(result_vividness$spearman_brown_ci95[2], 3), ")")),
c('Originality', paste0(round(result_originality$corr_median,3), " (", round(result_originality$corr_ci95[1], 3), " - ", round(result_originality$corr_ci95[2], 3), ")"),
paste0(round(result_originality$spearman_brown_median,3), " (", round(result_originality$spearman_brown_ci95[1], 3), " - ", round(result_originality$spearman_brown_ci95[2], 3), ")")),
c('Aesthetics', paste0(round(result_aesthetics$corr_median,3), " (", round(result_aesthetics$corr_ci95[1], 3), " - ", round(result_aesthetics$corr_ci95[2], 3), ")"),
paste0(round(result_aesthetics$spearman_brown_median,3), " (", round(result_aesthetics$spearman_brown_ci95[1], 3), " - ", round(result_aesthetics$spearman_brown_ci95[2], 3), ")")),
c('Curiosity', paste0(round(result_curiosity$corr_median,3), " (", round(result_curiosity$corr_ci95[1], 3), " - ", round(result_curiosity$corr_ci95[2], 3), ")"),
paste0(round(result_curiosity$spearman_brown_median,3), " (", round(result_curiosity$spearman_brown_ci95[1], 3), " - ", round(result_curiosity$spearman_brown_ci95[2], 3), ")")))

```


In this case, the Spearman-Brown corrected reliability of all 5 subscales oscillates between 0.77 and 0.86, indicating good to strong reliability. 



## Comment 6


```{r data for the correlations between raters}

corr_data <- aggregate(cbind(creativity, liking, vividness, originality, aesthetics, curiosity) ~ imageId + raterType + category, data=bd, FUN=mean)

corr_data <- corr_data %>% 
   pivot_wider(names_from = raterType, values_from = c(creativity, liking, vividness, originality, aesthetics, curiosity))

corr_data_allcat <- aggregate(cbind(creativity, liking, vividness, originality, aesthetics, curiosity) ~ imageId + raterType , data=bd, FUN=mean)

corr_data_allcat_long <- corr_data_allcat |> pivot_longer(
  cols = creativity:curiosity,
  names_to = 'scale',
  values_to = 'score'
)

corr_data_allcat_long_wide <- corr_data_allcat_long %>% 
   pivot_wider(names_from = raterType, values_from = score)

corr_data_allcat_long_wide$scale <- factor(corr_data_allcat_long_wide$scale, levels=c('creativity', 'liking', 'vividness', 'originality', 'aesthetics', 'curiosity'), labels=c('Creativity', 'Liking', 'Vividness', 'Originality', 'Aesthetics', 'Curiosity'))



(correlationPlotGpt <- ggplot(corr_data_allcat_long_wide, aes(y=Human, x=GPT)) + geom_point() + facet_wrap(~scale, nrow=1) + geom_smooth(method="lm", color='red') + 
  theme_classic() + stat_cor(aes(label = paste(..rr.label.., ..p.label.., sep = "*`,`~")), color = "red", geom = "label", p.accuracy=0.001, size=3) + 
  theme(legend.position = 'none', text=element_text(family ='serif'))+ ylab('Human score') + xlab('GPT Score') + ggtitle(''))

(correlationPlotGuidedGPT <- ggplot(corr_data_allcat_long_wide, aes(y=Human, x=GuidedGPT)) + geom_point() + facet_wrap(~scale, nrow=1) + geom_smooth(method="lm", color='red') + 
  theme_classic() + stat_cor(aes(label = paste(..rr.label.., ..p.label.., sep = "*`,`~")), color = "red", geom = "label", p.accuracy=0.001, size=3) + 
  theme(legend.position = 'none', text=element_text(family ='serif'))+ ylab('Human score') + xlab('Guided GPT Score') + ggtitle(''))

```



# Main analysis

## Description of the data


## Effect of category on the ratings


```{r modelling the effect of category on each scale for every type of rater}

raterTypes <- levels(bd$raterType)

modelsCategory <- list()
modelsCategoryNull <- list()
modelsCategorySummaries <- list()
categoryEffect <- list()

for (r in raterTypes) {

  modelsCategory[[r]] <- list()

  for (i in seq_along(scaleVariables)) {

    formulaModel <- as.formula(
      paste0(scaleVariables[i], " ~ category + (1|imageId)")
    )

    modelsCategory[[r]][[scaleVariables[i]]] <- lmer(
      formulaModel,
      data = bd[bd$raterType == r, ]
    )
    
    formulaModelNull <- as.formula(
      paste0(scaleVariables[i], " ~  (1|imageId)")
    )
    
    modelsCategoryNull[[r]][[scaleVariables[i]]] <- lmer(
      formulaModelNull,
      data = bd[bd$raterType == r, ]
    )
    
    model <- modelsCategory[[r]][[scaleVariables[i]]]
    
    modelsCategorySummaries[[r]][[scaleVariables[i]]] <- rbind(c('Fixed effects', '', '', '', '', ''),
                                 c('', 'Est/Beta', 'SE', '95% CI', 't', 'p value'),
                                 c('Group', '', '', '', '', ''),
                                 cbind(c('Intercept', levels(bd$category)[2:4]),
                                       round(summary(model)$coefficients[,1:2], 2),
                                       apply( round(confint(model, method='Wald')[3:6,], 2), 1 , paste , collapse = " - " ),
                                       round(summary(model)$coefficients[,4], 2),
                                       round(summary(model)$coefficients[,5], 3)),
                                 c('', '', '', '', '', ''),
                                 c('Random effects', '', '', '', '', ''),
                                 c('', '', '', '', 'Variance', 'S.D.'),
                                 c('Image (Intercept)', '', '', '', round(as.data.frame(VarCorr(model))[1,4], 3), round(as.data.frame(VarCorr(model))[1,5], 3)),
                                 c('', '', '', '', '', ''),
                                 c('Model fit', '', '', '', '', ''),
                                 c('', '', '', '', 'Marginal', 'Conditional'),
                                 c('R2', '', '', '', round(performance::r2(model)$R2_marginal,2), round(performance::r2(model)$R2_conditional, 2)))
    
    # In order to obtain a value for the effect of each variable in the model, I generate a series
    # of new models (I call them "Null") that miss one of the regressors each. This way, I can compare
    # the full model with each of the Nulls and obtain the effect of each of the regressors. As
    # results, I will report the Chi2, df, p.value, and difference in AIC.
    
    modelComparison <- anova(modelsCategory[[r]][[scaleVariables[i]]], modelsCategoryNull[[r]][[scaleVariables[i]]])
    categoryEffect[[r]][[scaleVariables[i]]] <- paste0('Chi(', modelComparison$Df[2], ')= ', 
                         round(modelComparison[2,'Chisq'], 2), 
                         ', p.value = ', round(modelComparison[2,'Pr(>Chisq)'], 3), 
                         ', Difference AIC = ', round(diff(modelComparison$AIC), 2))
    
    
  }
}


```


```{r values for the figures representing the ratings by category for each rater type}
summaries <- list()
individualMeans <- list()

summariesBd <- c()
individualMeansBd <- c()

for (i in 1:length(scaleVariables)) {
  
  imageMean <- aggregate(as.formula(paste(scaleVariables[i], ' ~ category + imageId + raterType')), data = bd, FUN = mean)
  
  summaries[[i]] <- aggregate(as.formula(paste(scaleVariables[i], ' ~ category + raterType')), data = imageMean, FUN = function(x) c(mn = mean(x), sd = sd(x), ln=length(x) ))
  
  summaries[[i]]$sd <- summaries[[i]][,3][,2]
  summaries[[i]]$len <- summaries[[i]][,3][,3]
  summaries[[i]]$score <- summaries[[i]][,3][,1]
  summaries[[i]]$sem <- summaries[[i]]$sd / sqrt(summaries[[i]]$len)
  summaries[[i]]$ymin <- summaries[[i]]$score - summaries[[i]]$sd
  summaries[[i]]$ymax <- summaries[[i]]$score + summaries[[i]]$sd
  summaries[[i]]$category <- factor(summaries[[i]]$category)
  summaries[[i]]$raterType <- factor(summaries[[i]]$raterType)
  
  summaries[[i]]$scale <- scaleVariables[i]
  summaries[[i]] <- summaries[[i]][,-3]
  individualMeans[[i]] <- aggregate(as.formula(paste(scaleVariables[i], ' ~ imageId + category + raterType')), data = bd, FUN=mean)
  names(individualMeans[[i]])[4] <- 'score'
  individualMeans[[i]][5] <- scaleVariables[i]
  names(individualMeans[[i]])[5] <- 'scale'
  
  summariesBd <- rbind(summariesBd, summaries[[i]])
  individualMeansBd <- rbind(individualMeansBd, individualMeans[[i]])
  
}

# summariesBd <- subset(summariesBd, summariesBd$scale %in% c('liking', 'vividness', 'originality', 'aesthetics', 'curiosity'))
# individualMeansBd <- subset(individualMeansBd, individualMeansBd$scale %in% c('liking', 'vividness', 'originality', 'aesthetics', 'curiosity'))

summariesBd$scale <- factor(summariesBd$scale, levels=c('liking', 'vividness', 'originality', 'aesthetics', 'curiosity', 'creativity'))
individualMeansBd$scale <- factor(individualMeansBd$scale, levels=c('liking', 'vividness', 'originality', 'aesthetics', 'curiosity', 'creativity'))



summariesBd$scale <- factor(summariesBd$scale, 
                            levels=c('liking', 'vividness', 'originality', 'aesthetics', 'curiosity', 'creativity'),
                            labels=c('Liking', 'Vividness', 'Originality', 'Aesthetics', 'Curiosity', 'creativity'))

individualMeansBd$scale <- factor(individualMeansBd$scale, 
                            levels=c('liking', 'vividness', 'originality', 'aesthetics', 'curiosity', 'creativity'),
                            labels=c('Liking', 'Vividness', 'Originality', 'Aesthetics', 'Curiosity', 'creativity'))


```


### For human raters


```{r results from the human rater models}

# Pairwise comparisons, mainly for the figures

round(summary(pairs(emmeans(modelsCategory$Human$creativity, pairwise ~ category)))$p.value, 3)[c(1,4,6)]

annotation_creativity_human <- c('***', 'n.s.', '***')

rbind(round(summary(pairs(emmeans(modelsCategory$Human$liking, pairwise ~ category)))$p.value, 3)[c(1,4,6)],
      round(summary(pairs(emmeans(modelsCategory$Human$vividness, pairwise ~ category)))$p.value, 3)[c(1,4,6)],
      round(summary(pairs(emmeans(modelsCategory$Human$originality, pairwise ~ category)))$p.value, 3)[c(1,4,6)],
      round(summary(pairs(emmeans(modelsCategory$Human$aesthetics, pairwise ~ category)))$p.value, 3)[c(1,4,6)],
      round(summary(pairs(emmeans(modelsCategory$Human$curiosity, pairwise ~ category)))$p.value, 3)[c(1,4,6)])

annotation_df_human <- data.frame(
  scale = rep(levels(summariesBd$scale)[1:5], each=3),       
  xmin = rep(c(1,2,3), times=5), xmax = rep(c(2,3,4), times=5),     
  y_position = rep(c(7.5, 7, 6.5), 5),    
  annotation = c('***', '**', '***',
                 '***', '*', '***',
                 '***', 'n.s.', '***',
                 '***', 'n.s.', '***',
                 '***', 'n.s.', '***')  
)


# Model details for the text
categoryEffect$Human


modelsCategorySummaries$Human$creativity
```


```{r}
modelHumanCategory2 <- lmer(creativity ~ category2 + (1|imageId), data=bd[bd$raterType=='Human',])
modelHumanCategory2Null  <- lmer(creativity ~  (1|imageId), data=bd[bd$raterType=='Human',])

modelComparison <- anova(modelHumanCategory2, modelHumanCategory2Null)
paste0('Chi(', modelComparison$Df[2], ')= ', 
        round(modelComparison[2,'Chisq'], 2), 
        ', p.value = ', round(modelComparison[2,'Pr(>Chisq)'], 3), 
        ', Difference AIC = ', round(diff(modelComparison$AIC), 2))

emmeans(modelHumanCategory2, pairwise ~  category2)



modelCategory2HumansSummary <- rbind(c('Fixed effects', '', '', '', '', ''),
                                 c('', 'Est/Beta', 'SE', '95% CI', 't', 'p value'),
                                 c('Group', '', '', '', '', ''),
                                 cbind(c(rownames(summary(modelHumanCategory2)$coefficients)),
                                       round(summary(modelHumanCategory2)$coefficients[,1:2], 2),
                                       apply( round(confint(modelHumanCategory2, method='Wald')[3:4,], 2), 1 , paste , collapse = " - " ),
                                       round(summary(modelHumanCategory2)$coefficients[,4], 2),
                                       round(summary(modelHumanCategory2)$coefficients[,5], 3)),
                                 c('', '', '', '', '', ''),
                                 c('Random effects', '', '', '', '', ''),
                                 c('', '', '', '', 'Variance', 'S.D.'),
                                 c('Image (Intercept)', '', '', '', round(as.data.frame(VarCorr(modelHumanCategory2))[1,4], 3), round(as.data.frame(VarCorr(modelHumanCategory2))[1,5], 3)),
                                 c('', '', '', '', '', ''),
                                 c('Model fit', '', '', '', '', ''),
                                 c('', '', '', '', 'Marginal', 'Conditional'),
                                 c('R2', '', '', '', round(performance::r2(modelHumanCategory2)$R2_marginal,2), round(performance::r2(modelHumanCategory2)$R2_conditional, 2)))



```


```{r}
bd$categoryNonArtists <- factor(bd$category, levels(bd$category)[c(2,1,3,4)])

modelHumanCategoryNonArtists <- lmer(creativity ~ categoryNonArtists + (1|imageId), data=bd[bd$raterType=='Human',])



summary(modelHumanCategoryNonArtists)
```


```{r}


(violinPlotHuman <- ggplot(data = individualMeansBd[(individualMeansBd$raterType=='Human' & individualMeansBd$scale=='creativity'),], aes(x=category, y=score, fill=category, colour=category)) + 
  geom_violin(alpha=0.4) + scale_y_continuous(limits = c(-8, 8), breaks=c(0, 2, 4, 6, 8)) +
  theme_classic() +
  scale_fill_manual(values=selected_colours) + 
  scale_colour_manual(values=selected_colours) +
  ggnewscale::new_scale_color() +
  ggnewscale::new_scale_fill() +
  geom_point(data=individualMeans[[6]][individualMeans[[6]]$raterType=='Human',], aes(y=score, x=category, colour = category), 
             position = position_jitter(w = 0.35, h = 0), alpha=0.6, size=0.9)+ 
  scale_colour_manual(values=selected_colours) + 
  scale_fill_manual(values=selected_colours) + 
  theme(legend.position='none', text=element_text(family ='serif'), axis.text.x = element_blank())+ 
  labs(x = '', y ='Mean creativity score') + scale_x_discrete(
    labels=categoryLabels)+
  ylim(c(1, 7)) +
  geom_signif(annotations = annotation_creativity_human, y_position =c(6.5, 6, 5.5),xmin = c(1,2,3), xmax = c(2,3,4),tip_length = 0.05, family = 'serif') + 
  geom_errorbar(data = summariesBd[summariesBd$raterType=='Human' & summariesBd$scale=='creativity',], 
                aes(x = category, ymin = ymin, ymax = ymax), inherit.aes=FALSE, 
                width = 0.2, color = "black", size = 1) +
  geom_point(data = summariesBd[summariesBd$raterType=='Human' & summariesBd$scale=='creativity',], aes(x = category, y = score), inherit.aes=FALSE,
             color = "black",  shape = 16))









annotation_df_human$scale <- factor(annotation_df_human$scale, levels=c('Liking', 'Vividness', 'Originality', 'Aesthetics', 'Curiosity'))



(plotBarsScalesHuman <-  ggplot() + 
  geom_bar(data=summariesBd[summariesBd$raterType=='Human' & summariesBd$scale!='creativity',], aes(y=score, x=category, fill=category, color=category), 
           stat="identity", position=position_dodge(), alpha=0.4)  +
  scale_fill_manual(values=selected_colours) + 
  scale_colour_manual(values=selected_colours) +
  ggnewscale::new_scale_color() +
  geom_point(data=individualMeansBd[individualMeansBd$raterType=='Human' & individualMeansBd$scale!='creativity',], aes(y=score, x=category, colour = category), 
             position = position_jitter(w = 0.35, h = 0), size = 0.3) +
  geom_errorbar(data=summariesBd[summariesBd$raterType=='Human' & summariesBd$scale!='creativity',],aes( x=category, ymin=ymin, ymax=ymax), width=.2,
                position=position_dodge(.9)) +
  scale_colour_manual(values=selected_colours) + 
  theme_classic() + scale_y_continuous(expand = c(0, 0)) + 
  theme(legend.position = 'none', axis.text.x = element_blank(), text = element_text(family = "serif"), strip.text.x = element_text(), strip.background = element_blank() )+ 
  labs(x = '', y ='Mean score', title='Humans') + coord_cartesian(ylim = c(0,8)) + facet_grid(~ scale)+
  geom_signif(
    data = annotation_df_human,
    aes(xmin = xmin, xmax = xmax, annotations = annotation, y_position = y_position,  group = interaction(xmin, xmax, scale)),
    manual = TRUE,
    tip_length = 0.02, 
    family='serif'
  ))



```



### For GPT ratings


```{r results from the GPT rater models}

# Pairwise comparisons, mainly for the figures

round(summary(pairs(emmeans(modelsCategory$GPT$creativity, pairwise ~ category)))$p.value, 3)[c(1,4,6)]

annotation_creativity_gpt <- c('n.s', 'n.s.', '***')

rbind(round(summary(pairs(emmeans(modelsCategory$GPT$liking, pairwise ~ category)))$p.value, 3)[c(1,4,6)],
      round(summary(pairs(emmeans(modelsCategory$GPT$vividness, pairwise ~ category)))$p.value, 3)[c(1,4,6)],
      round(summary(pairs(emmeans(modelsCategory$GPT$originality, pairwise ~ category)))$p.value, 3)[c(1,4,6)],
      round(summary(pairs(emmeans(modelsCategory$GPT$aesthetics, pairwise ~ category)))$p.value, 3)[c(1,4,6)],
      round(summary(pairs(emmeans(modelsCategory$GPT$curiosity, pairwise ~ category)))$p.value, 3)[c(1,4,6)])


annotation_df_gpt <- data.frame(
  scale = rep(levels(summariesBd$scale)[1:5], each=3),       
  xmin = rep(c(1,2,3), times=5), xmax = rep(c(2,3,4), times=5),     
  y_position = rep(c(7.5, 7, 6.5), 5),    
  annotation = c('n.s.', 'n.s.', '***',
                 'n.s.', '*', '***',
                 'n.s.', '**', '**',
                 'n.s.', 'n.s.', '***',
                 'n.s.', '*', '***')  
)

# Model details for the text
categoryEffect$GPT

```

```{r}
modelGptCategory2 <- lmer(creativity ~ category2 + (1|imageId), data=bd[bd$raterType=='GPT',])
modelGptCategory2Null  <- lmer(creativity ~  (1|imageId), data=bd[bd$raterType=='GPT',])

modelComparison <- anova(modelGptCategory2, modelGptCategory2Null)
paste0('Chi(', modelComparison$Df[2], ')= ', 
        round(modelComparison[2,'Chisq'], 2), 
        ', p.value = ', round(modelComparison[2,'Pr(>Chisq)'], 3), 
        ', Difference AIC = ', round(diff(modelComparison$AIC), 2))

emmeans(modelGptCategory2, pairwise ~  category2)



rbind(c('Fixed effects', '', '', '', '', ''),
                                 c('', 'Est/Beta', 'SE', '95% CI', 't', 'p value'),
                                 c('Group', '', '', '', '', ''),
                                 cbind(c(rownames(summary(modelGptCategory2)$coefficients)),
                                       round(summary(modelGptCategory2)$coefficients[,1:2], 2),
                                       apply( round(confint(modelGptCategory2, method='Wald')[3:4,], 2), 1 , paste , collapse = " - " ),
                                       round(summary(modelGptCategory2)$coefficients[,4], 2),
                                       round(summary(modelGptCategory2)$coefficients[,5], 3)),
                                 c('', '', '', '', '', ''),
                                 c('Random effects', '', '', '', '', ''),
                                 c('', '', '', '', 'Variance', 'S.D.'),
                                 c('Image (Intercept)', '', '', '', round(as.data.frame(VarCorr(modelGptCategory2))[1,4], 3), round(as.data.frame(VarCorr(modelGptCategory2))[1,5], 3)),
                                 c('', '', '', '', '', ''),
                                 c('Model fit', '', '', '', '', ''),
                                 c('', '', '', '', 'Marginal', 'Conditional'),
                                 c('R2', '', '', '', round(performance::r2(modelGptCategory2)$R2_marginal,2), round(performance::r2(modelGptCategory2)$R2_conditional, 2)))



```



```{r}

(violinPlotGPT <- ggplot(data = individualMeansBd[(individualMeansBd$raterType=='GPT' & individualMeansBd$scale=='creativity'),], aes(x=category, y=score, fill=category, colour=category)) + 
  geom_violin(alpha=0.4) + scale_y_continuous(limits = c(-8, 8), breaks=c(0, 2, 4, 6, 8)) +
  theme_classic() +
  scale_fill_manual(values=selected_colours) + 
  scale_colour_manual(values=selected_colours) +
  ggnewscale::new_scale_color() +
  ggnewscale::new_scale_fill() +
  geom_point(data=individualMeans[[6]][individualMeans[[6]]$raterType=='GPT',], aes(y=score, x=category, colour = category), 
             position = position_jitter(w = 0.35, h = 0), alpha=0.6, size=0.9)+ 
  scale_colour_manual(values=selected_colours) + 
  scale_fill_manual(values=selected_colours) + 
  theme(legend.position='none', text=element_text(family ='serif'), axis.text.x = element_blank())+ 
  labs(x = '', y ='Mean creativity score') + scale_x_discrete(
    labels=categoryLabels)+
  ylim(c(1, 7)) +
  geom_signif(annotations = annotation_creativity_gpt, y_position =c(6.5, 6, 5.5),xmin = c(1,2,3), xmax = c(2,3,4),tip_length = 0.05, family = 'serif') + 
  geom_errorbar(data = summariesBd[summariesBd$raterType=='GPT' & summariesBd$scale=='creativity',], 
                aes(x = category, ymin = ymin, ymax = ymax), inherit.aes=FALSE, 
                width = 0.2, color = "black", size = 1) +
  geom_point(data = summariesBd[summariesBd$raterType=='GPT' & summariesBd$scale=='creativity',], aes(x = category, y = score), inherit.aes=FALSE,
             color = "black",  shape = 16))






annotation_df_gpt$scale <- factor(annotation_df_gpt$scale, levels=c('Liking', 'Vividness', 'Originality', 'Aesthetics', 'Curiosity'))

 
 
(plotBarsScalesGPT <-  ggplot() + 
  geom_bar(data=summariesBd[summariesBd$raterType=='GPT' & summariesBd$scale!='creativity',], aes(y=score, x=category, fill=category, color=category), 
           stat="identity", position=position_dodge(), alpha=0.4)  +
  scale_fill_manual(values=selected_colours) + 
  scale_colour_manual(values=selected_colours) +
  ggnewscale::new_scale_color() +
  geom_point(data=individualMeansBd[individualMeansBd$raterType=='GPT' & individualMeansBd$scale!='creativity',], aes(y=score, x=category, colour = category), 
             position = position_jitter(w = 0.35, h = 0), size = 0.3) +
  geom_errorbar(data=summariesBd[summariesBd$raterType=='GPT' & summariesBd$scale!='creativity',],aes( x=category, ymin=ymin, ymax=ymax), width=.2,
                position=position_dodge(.9)) +
  scale_colour_manual(values=selected_colours) + 
  theme_classic() + scale_y_continuous(expand = c(0, 0)) + 
  theme(legend.position = 'none', axis.text.x = element_blank(), text = element_text(family = "serif"), strip.text.x = element_text(), strip.background = element_blank() )+ 
  labs(x = '', y ='Mean score', title='GPT') + coord_cartesian(ylim = c(0, 8)) + facet_grid(~ scale)+
  geom_signif(
    data = annotation_df_gpt,
    aes(xmin = xmin, xmax = xmax, annotations = annotation, y_position = y_position,  group = interaction(xmin, xmax, scale)),
    manual = TRUE,
    tip_length = 0.02, 
    family='serif'
  ))
 



```



### For Guided GPT ratings


```{r results from the Guided GPT rater models}

# Pairwise comparisons, mainly for the figures

round(summary(pairs(emmeans(modelsCategory$GuidedGPT$creativity, pairwise ~ category)))$p.value, 3)[c(1,4,6)]

annotation_creativity_guidedgpt <- c('***', 'n.s.', '***')

rbind(round(summary(pairs(emmeans(modelsCategory$GuidedGPT$liking, pairwise ~ category)))$p.value, 3)[c(1,4,6)],
      round(summary(pairs(emmeans(modelsCategory$GuidedGPT$vividness, pairwise ~ category)))$p.value, 3)[c(1,4,6)],
      round(summary(pairs(emmeans(modelsCategory$GuidedGPT$originality, pairwise ~ category)))$p.value, 3)[c(1,4,6)],
      round(summary(pairs(emmeans(modelsCategory$GuidedGPT$aesthetics, pairwise ~ category)))$p.value, 3)[c(1,4,6)],
      round(summary(pairs(emmeans(modelsCategory$GuidedGPT$curiosity, pairwise ~ category)))$p.value, 3)[c(1,4,6)])


annotation_df_guidedGPT <- data.frame(
  scale = rep(levels(summariesBd$scale)[1:5], each=3),       
  xmin = rep(c(1,2,3), times=5), xmax = rep(c(2,3,4), times=5),     
  y_position = rep(c(7.5, 7, 6.5), 5),    
  annotation = c('**', '***', '***',
                 '*', '***', '***',
                 '**', '***', '***',
                 '*', '***', '***',
                 '**', '***', '***')  
)

# Model details for the text
categoryEffect$GuidedGPT

```
```{r}

(violinPlotGuidedGPT <- ggplot(data = individualMeansBd[(individualMeansBd$raterType=='GuidedGPT' & individualMeansBd$scale=='creativity'),], aes(x=category, y=score, fill=category, colour=category)) + 
  geom_violin(alpha=0.4) + scale_y_continuous(limits = c(-8, 8), breaks=c(0, 2, 4, 6, 8)) +
  theme_classic() +
  scale_fill_manual(values=selected_colours) + 
  scale_colour_manual(values=selected_colours) +
  ggnewscale::new_scale_color() +
  ggnewscale::new_scale_fill() +
  geom_point(data=individualMeans[[6]][individualMeans[[6]]$raterType=='GuidedGPT',], aes(y=score, x=category, colour = category), 
             position = position_jitter(w = 0.35, h = 0), alpha=0.6, size=0.9)+ 
  scale_colour_manual(values=selected_colours) + 
  scale_fill_manual(values=selected_colours) + 
  theme(legend.position='none', text=element_text(family ='serif'), axis.text.x = element_blank())+ 
  labs(x = '', y ='Mean creativity score') + scale_x_discrete(
    labels=categoryLabels)+
  ylim(c(1, 7)) +
  geom_signif(annotations = annotation_creativity_guidedgpt, y_position =c(6.5, 6, 5.5),xmin = c(1,2,3), xmax = c(2,3,4),tip_length = 0.05, family = 'serif') + 
  geom_errorbar(data = summariesBd[summariesBd$raterType=='GuidedGPT' & summariesBd$scale=='creativity',], 
                aes(x = category, ymin = ymin, ymax = ymax), inherit.aes=FALSE, 
                width = 0.2, color = "black", size = 1) +
  geom_point(data = summariesBd[summariesBd$raterType=='GuidedGPT' & summariesBd$scale=='creativity',], aes(x = category, y = score), inherit.aes=FALSE,
             color = "black",  shape = 16))







annotation_df_guidedGPT$scale <- factor(annotation_df_guidedGPT$scale, levels=c('Liking', 'Vividness', 'Originality', 'Aesthetics', 'Curiosity'))

  
(plotBarsScalesGuidedGPT <-  ggplot() + 
  geom_bar(data=summariesBd[summariesBd$raterType=='GuidedGPT' & summariesBd$scale!='creativity',], aes(y=score, x=category, fill=category, color=category), 
           stat="identity", position=position_dodge(), alpha=0.4)  + 
  scale_fill_manual(values=selected_colours) + 
  scale_colour_manual(values=selected_colours) +
  ggnewscale::new_scale_color() +
  geom_point(data=individualMeansBd[individualMeansBd$raterType=='GuidedGPT' & individualMeansBd$scale!='creativity',], aes(y=score, x=category, colour = category), 
             position = position_jitter(w = 0.35, h = 0), size = 0.3) +
  geom_errorbar(data=summariesBd[summariesBd$raterType=='GuidedGPT' & summariesBd$scale!='creativity',],aes( x=category, ymin=ymin, ymax=ymax), width=.2,
                position=position_dodge(.9)) +
  scale_colour_manual(values=selected_colours) + 
  theme_classic() + scale_y_continuous() + 
  theme(legend.position = 'none', axis.text.x = element_blank(), text = element_text(family = "serif"), strip.text.x = element_text(), strip.background = element_blank() )+ 
  labs(x = '', y ='Mean score', title='Guided GPT') + coord_cartesian(ylim = c(0, 8)) + facet_grid(~ scale)+
  geom_signif(
    data = annotation_df_guidedGPT,
    aes(xmin = xmin, xmax = xmax, annotations = annotation, y_position = y_position,  group = interaction(xmin, xmax, scale)),
    manual = TRUE,
    tip_length = 0.02, 
    family='serif'
  ))


```



```{r arrange figure 3 for the paper}

legendPlot <-  ggplot() + 
  geom_bar(data=summariesBd[summariesBd$raterType=='GuidedGPT' & summariesBd$scale!='creativity',], aes(y=score, x=category, fill=category, color=category), 
           stat="identity", position=position_dodge(), alpha=0.4)  +
  scale_fill_manual(values=selected_colours, name = "Category") + 
  scale_colour_manual(values=selected_colours, name = "Category") +
  ggnewscale::new_scale_color() +
  geom_point(data=individualMeansBd[individualMeansBd$raterType=='GuidedGPT' & individualMeansBd$scale!='creativity',], aes(y=score, x=category, colour = category), 
             position = position_jitter(w = 0.35, h = 0), size = 0.3) +
  geom_errorbar(data=summariesBd[summariesBd$raterType=='GuidedGPT' & summariesBd$scale!='creativity',],aes( x=category, ymin=ymin, ymax=ymax), width=.2,
                position=position_dodge(.9)) +
  scale_colour_manual(values=selected_colours, name = "Category") + 
  theme_classic() + scale_y_continuous() + 
  theme(legend.position = 'bottom', axis.text.x = element_blank(), text = element_text(family = "serif"), strip.text.x = element_text(), strip.background = element_blank() )+ 
  labs(x = '', y ='Mean score', title='') + coord_cartesian(ylim = c(0, 8)) + facet_grid(~ scale)+
  geom_signif(
    data = annotation_df_guidedGPT,
    aes(xmin = xmin, xmax = xmax, annotations = annotation, y_position = y_position,  group = interaction(xmin, xmax, scale)),
    manual = TRUE,
    tip_length = 0.02, 
    family='serif'
  )



ggarrange(
ggarrange(violinPlotHuman, plotBarsScalesHuman, nrow=1, widths=c(1.5, 5), labels=c('A', 'B')),
ggarrange(violinPlotGPT, plotBarsScalesGPT, nrow=1, widths=c(1.5, 5), labels=c('C', 'D')),
ggarrange(violinPlotGuidedGPT, plotBarsScalesGuidedGPT, nrow=1, widths=c(1.5, 5), labels=c('E', 'F')), 
get_legend(legendPlot), nrow=4, heights=c(5, 5, 5, 1))


```


## Differences in ratings between rater types

```{r}
modelRaterCat <- lmer(creativity ~ raterType * category + (1|imageId), data = bd)

summary(modelRaterCat)



modelRaterCategorySummary <- rbind(c('Fixed effects', '', '', '', '', ''),
                                 c('', 'Est/Beta', 'SE', '95% CI', 't', 'p value'),
                                 c('Group', '', '', '', '', ''),
                                 cbind(c(rownames(summary(modelRaterCat)$coefficients)),
                                       round(summary(modelRaterCat)$coefficients[,1:2], 2),
                                       apply( round(confint(modelRaterCat, method='Wald')[3:14,], 2), 1 , paste , collapse = " - " ),
                                       round(summary(modelRaterCat)$coefficients[,4], 2),
                                       round(summary(modelRaterCat)$coefficients[,5], 3)),
                                 c('', '', '', '', '', ''),
                                 c('Random effects', '', '', '', '', ''),
                                 c('', '', '', '', 'Variance', 'S.D.'),
                                 c('Image (Intercept)', '', '', '', round(as.data.frame(VarCorr(modelRaterCat))[1,4], 3), round(as.data.frame(VarCorr(modelRaterCat))[1,5], 3)),
                                 c('', '', '', '', '', ''),
                                 c('Model fit', '', '', '', '', ''),
                                 c('', '', '', '', 'Marginal', 'Conditional'),
                                 c('R2', '', '', '', round(performance::r2(modelRaterCat)$R2_marginal,2), round(performance::r2(modelRaterCat)$R2_conditional, 2)))



emmeans(modelRaterCat, pairwise ~ raterType | category)



modelRaterCatNullInteraction <- lmer(creativity ~ raterType + category + (1|imageId), data = bd)
modelRaterCatNullRater <- lmer(creativity ~  category + (1|imageId), data = bd)
modelRaterCatNullCategory <- lmer(creativity ~ raterType + (1|imageId), data = bd)


modelComparison <- anova(modelRaterCat, modelRaterCatNullCategory)
paste0('Chi(', modelComparison$Df[2], ')= ', 
        round(modelComparison[2,'Chisq'], 2), 
        ', p.value = ', round(modelComparison[2,'Pr(>Chisq)'], 3), 
        ', Difference AIC = ', round(diff(modelComparison$AIC), 2))



modelComparison <- anova(modelRaterCat, modelRaterCatNullRater)
paste0('Chi(', modelComparison$Df[2], ')= ', 
        round(modelComparison[2,'Chisq'], 2), 
        ', p.value = ', round(modelComparison[2,'Pr(>Chisq)'], 3), 
        ', Difference AIC = ', round(diff(modelComparison$AIC), 2))


modelComparison <- anova(modelRaterCat, modelRaterCatNullInteraction)
paste0('Chi(', modelComparison$Df[2], ')= ', 
        round(modelComparison[2,'Chisq'], 2), 
        ', p.value = ', round(modelComparison[2,'Pr(>Chisq)'], 3), 
        ', Difference AIC = ', round(diff(modelComparison$AIC), 2))



```


```{r figure effect of rater type and category on creativity}



bdAgg <- aggregate(creativity ~ imageId + raterType + category, data=bd, FUN=mean)




(plotRaterCategory <- ggplot(bdAgg, aes(x = category, y = creativity, fill = raterType)) +
  geom_boxplot(outlier.shape=NA, alpha=0.6) +
  theme_classic(base_size = 14) + 
  theme(axis.ticks.x = element_blank(), text=element_text(family ='serif'), legend.position='bottom') + 
  scale_fill_manual(values=coloursRaters, name = 'Rater type')+
  ylim(c(1, 7)) +
  geom_signif(annotations = c('***', 'n.s.', '***', '***'), 
              y_position =c(5.5, 5.5, 5.5, 5.5),
              xmin = c(0.7,1.7,2.7,3.7), 
              xmax = c(0.95,1.95, 2.95,3.95),tip_length = 0.01,
              textsize = 4, family='serif') +
      geom_signif(annotations = c('***', '***', '***', '***'), 
              y_position =c(5.5, 5.5, 5.5, 5.5),
              xmin = c(1.05,2.05, 3.05,4.05), 
              xmax = c(1.3,2.3, 3.3,4.3),tip_length = 0.01,
              textsize = 4, family='serif') +
      geom_signif(annotations = c('**', '***', '***', '***'), 
              y_position =c(6, 6, 6, 6),
              xmin = c(0.7,1.7,2.7,3.7), 
              xmax = c(1.3, 2.3, 3.3, 4.3),tip_length = 0.01,
              textsize = 4, family='serif') +
  ylab('Creativity score') + xlab('Category'))

```

```{r arrange figure 4}
ggarrange(correlationPlotGpt,
          correlationPlotGuidedGPT,
          plotRaterCategory,
          nrow=3, labels = c('A', 'B', 'C'), heights=c(1,1,1.5))
```



## Effect of AReA on ratings



```{r plot area groups}
humans <- merge(humans, dm, by='participantId', all.x=T)
modelAreaCat <- lmer(creativity ~ areaGroup * category + (1|imageId), data=humans)
emmeans(modelAreaCat, pairwise ~  areaGroup)



modelAreaCategorySummary <- rbind(c('Fixed effects', '', '', '', '', ''),
                                 c('', 'Est/Beta', 'SE', '95% CI', 't', 'p value'),
                                 c('Group', '', '', '', '', ''),
                                 cbind(c(rownames(summary(modelAreaCat)$coefficients)),
                                       round(summary(modelAreaCat)$coefficients[,1:2], 2),
                                       apply( round(confint(modelAreaCat, method='Wald')[3:10,], 2), 1 , paste , collapse = " - " ),
                                       round(summary(modelAreaCat)$coefficients[,4], 2),
                                       round(summary(modelAreaCat)$coefficients[,5], 3)),
                                 c('', '', '', '', '', ''),
                                 c('Random effects', '', '', '', '', ''),
                                 c('', '', '', '', 'Variance', 'S.D.'),
                                 c('Image (Intercept)', '', '', '', round(as.data.frame(VarCorr(modelAreaCat))[1,4], 3), round(as.data.frame(VarCorr(modelAreaCat))[1,5], 3)),
                                 c('', '', '', '', '', ''),
                                 c('Model fit', '', '', '', '', ''),
                                 c('', '', '', '', 'Marginal', 'Conditional'),
                                 c('R2', '', '', '', round(performance::r2(modelAreaCat)$R2_marginal,2), round(performance::r2(modelAreaCat)$R2_conditional, 2)))





modelAreaCatNullInteraction <- lmer(creativity ~ areaGroup + category + (1|imageId), data = humans)
modelAreaCatNullArea <- lmer(creativity ~  category + (1|imageId), data = humans)
modelAreaCatNullCategory <- lmer(creativity ~ areaGroup + (1|imageId), data = humans)


modelComparison <- anova(modelAreaCat, modelAreaCatNullCategory)
paste0('Chi(', modelComparison$Df[2], ')= ', 
        round(modelComparison[2,'Chisq'], 2), 
        ', p.value = ', round(modelComparison[2,'Pr(>Chisq)'], 3), 
        ', Difference AIC = ', round(diff(modelComparison$AIC), 2))



modelComparison <- anova(modelAreaCat, modelAreaCatNullArea)
paste0('Chi(', modelComparison$Df[2], ')= ', 
        round(modelComparison[2,'Chisq'], 2), 
        ', p.value = ', round(modelComparison[2,'Pr(>Chisq)'], 3), 
        ', Difference AIC = ', round(diff(modelComparison$AIC), 2))


modelComparison <- anova(modelAreaCat, modelAreaCatNullInteraction)
paste0('Chi(', modelComparison$Df[2], ')= ', 
        round(modelComparison[2,'Chisq'], 2), 
        ', p.value = ', round(modelComparison[2,'Pr(>Chisq)'], 3), 
        ', Difference AIC = ', round(diff(modelComparison$AIC), 2))

```


```{r}


bdAggAreaGroups <- aggregate(creativity ~ imageId + areaGroup + category, data=humans, FUN=mean)

(plotArea<-  ggplot(bdAggAreaGroups, aes(x = areaGroup, y = creativity, fill = category)) +
  geom_boxplot(outlier.shape=NA, alpha=0.6) +
  theme_classic(base_size = 16) +
  theme( text=element_text(family ='serif'), legend.position='none') +
  scale_fill_manual(values=selected_colours, name = 'Category')+
  ylim(c(1,7)) +
   geom_signif(annotations = c('***', 'n.s.', '***'), y_position =c(5.5, 5.25, 5),xmin = c(0.7, 0.9, 1.1), xmax = c(0.9, 1.1, 1.3),tip_length = 0.01,textsize = 6, family = 'serif') +
   geom_signif(annotations = c('***', 'n.s.', '***'), y_position =c(5.5, 5.25, 5),xmin = c(1.7, 1.9, 2.1), xmax = c(1.9, 2.1, 2.3),tip_length = 0.01,textsize = 6, family = 'serif') +
   geom_signif(annotations = c('', '', '***'), y_position =c(6.2, 6.2, 6.3),xmin = c(0.7, 1.7, 1), xmax = c(1.3, 2.3, 2),tip_length = 0.01,textsize =6, family = 'serif') +
   ylab('Creativity score') + xlab('AReA group'))



legendCategory <- ggplot(bd) + geom_bar(aes(x=category, fill=category)) + scale_fill_manual(values=selected_colours) + labs(fill='Category') + theme(legend.position = 'bottom')




```


```{r arrange figure 6}

ggarrange(plotArea, get_legend(legendCategory + theme(legend.position='bottom', legend.text = element_text(size=13, family='serif'), legend.title = element_text(size=13, family='serif'))),  heights=c(1, 0.09), nrow=2)

```



```{r save all summary tables into excel}

# by saving the summary tables to excel we will save time moving them to the manuscript in word



wb <- createWorkbook()

add_table <- function(wb, sheet, data) {
  addWorksheet(wb, sheet)
  writeData(wb, sheet, data)
}

add_table(wb, "Table S1", modelCategory2HumansSummary)
add_table(wb, "Table S2", modelsCategorySummaries$Human$creativity)
add_table(wb, "Table S3", modelsCategorySummaries$Human$liking)
add_table(wb, "Table S4", modelsCategorySummaries$Human$vividness)
add_table(wb, "Table S5", modelsCategorySummaries$Human$originality)
add_table(wb, "Table S6", modelsCategorySummaries$Human$aesthetics)
add_table(wb, "Table S7", modelsCategorySummaries$Human$curiosity)

add_table(wb, "Table S8", modelsCategorySummaries$GPT$creativity)
add_table(wb, "Table S9", modelsCategorySummaries$GuidedGPT$creativity)

add_table(wb, "Table S10", modelsCategorySummaries$GPT$liking)
add_table(wb, "Table S11", modelsCategorySummaries$GPT$vividness)
add_table(wb, "Table S12", modelsCategorySummaries$GPT$originality)
add_table(wb, "Table S13", modelsCategorySummaries$GPT$aesthetics)
add_table(wb, "Table S14", modelsCategorySummaries$GPT$curiosity)

add_table(wb, "Table S15", modelsCategorySummaries$GuidedGPT$liking)
add_table(wb, "Table S16", modelsCategorySummaries$GuidedGPT$vividness)
add_table(wb, "Table S17", modelsCategorySummaries$GuidedGPT$originality)
add_table(wb, "Table S18", modelsCategorySummaries$GuidedGPT$aesthetics)
add_table(wb, "Table S19", modelsCategorySummaries$GuidedGPT$curiosity)

add_table(wb, "Table S20", modelRaterCategorySummary)
add_table(wb, "Table S22", modelAreaCategorySummary)

saveWorkbook(wb, "results/tables/table-summaries.xlsx", overwrite = TRUE)
```

